Minimal trap spaces of Boolean networks are maximal siphons of their Petri net encoding

### 2023-01-02 manuscripts/TCS2022/tcs2022.tex
Finished the computation of special types of trap spaces

First, maximal trap spaces // TODO:

Second, we consider fixed points in Boolean networks.
Let \(s\) be a fixed point of a Boolean network \(\mathcal{N}\).
We have a subspace \(m\) corresponding to \(s\) as follows: \(\forall v \in V, m(v) = s(v)\), i.e., all nodes are fixed in \(m\).
Clearly, \(s\) is a trap set of \(\mathcal{N}\) regardless of the update scheme.
Hence, \(m\) is a trap space of \(\mathcal{N}\).
In addition, since \(\left\vert S_{\mathcal{N}}[m]\right\vert = 1\), \(m\) is also a minimal trap space.
To compute all fixed points of \(\mathcal{N}\), we can add more constraints to the encoded ASP characterizing all conflict-free siphons (equivalently trap spaces).
For every \(v \in V\), we add to the encoded ASP the rule
\[
  \verb|p-v| ; \verb|n-v|.
\]
that ensures that for every conflict-free siphon \(S\), it contains either \(\verb|p-v|\) or \(\verb|n-v|\) for every \(v \in V\).
Equivalently, the trap space corresponding to \(S\) is always a fixed point.
Now, the set of answer sets of the encoded ASP is equivalent to the set of fixed points of \(\mathcal{N}\).
In particular, when solving the encoded ASP using an ASP solver, we do not need to use the built-in option for computing set-inclusion maximal answer sets.
Note that we can also build another ASP characterizing all fixed points of \(\mathcal{N}\) based on the equivalence between a fixed point of \(\mathcal{N}\) and a deadlock of its Petri net encoding~\cite{chatain2014characterization}.
This approach may give a more compact ASP.

Third, trap spaces inside a subspace // TODO:

Finally, trap spaces that intersect a subspace // TODO:

>>>>

First, the work by~\cite{Rozum2021} uses the concept of stable motifs to build the succession diagram of a Boolean network, a summary of the decisions in the network dynamics that lead to successively more restrictive nested stable motifs.
The succession diagram is useful for control and decision making on this Boolean network.
In particular, the proposed control methods are independent to the update scheme.
It has been shown that a stable motif of a Boolean network is equivalent to a maximal trap space of this Boolean network~\cite{Rozum2021}.
Hence, it is necessary to develop an efficient method for computing maximal trap spaces of a Boolean network.
We shall show how to adjust the ASP-method presented in Subsection~\ref{subsec:computation_asp} to compute maximal trap spaces.

We first provide the definition of maximal trap spaces.
Let \(\varepsilon\) be the special trap space of \(\mathcal{N}\) where all the nodes are free.
Of course, \(\varepsilon\) corresponds to the special conflict-free siphon \(\emptyset\).
A trap space \(m\) is called maximal if \(m \neq \varepsilon\) and there is no other trap space \(m'\) such that \(m' \neq \varepsilon\) and \(m < m'\).
Analogously, a conflict-free siphon \(S\) is called minimal if \(S \neq \emptyset\) and there is no other trap space \(S'\) such that \(S' \neq \emptyset\) and \(S' \subset S\).
By using the reasoning similar to the proof of Theorem~\ref{theo:min_ts_2_max_sp}, we can easily conclude that a maximal trap space of \(\mathcal{N}\) is equivalent to a minimal conflict-free siphon of its Petri net \(\mathcal{P}\).
Let \(\mathcal{L}\) be the ASP characterizing all siphons of \(\mathcal{P}\) (see Subsection~\ref{subsec:computation_asp}).
Naturally, we need to exclude \(\emptyset\) from the solution space of \(\mathcal{L}\) (equivalently exclude \(\varepsilon\) from the set of trap spaces).
To do this, we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|p-v_1| ; \verb|n-v_1| ; \dots ; \verb|p-v_n| ; \verb|n-v_n|.
\]
that ensures that every answer set of \(\mathcal{L}\) cannot be empty.
Then a set-inclusion minimal answer set of \(\mathcal{L}\) is equivalent to a minimal conflict-free siphon of \(\mathcal{P}\), thus a maximal trap space of \(\mathcal{N}\)

Second, we consider fixed points in Boolean networks.
Let \(s\) be a fixed point of a Boolean network \(\mathcal{N}\).
We have a subspace \(m\) corresponding to \(s\) as follows: \(\forall v \in V, m(v) = s(v)\), i.e., all nodes are fixed in \(m\).
Clearly, \(s\) is a trap set of \(\mathcal{N}\) regardless of the update scheme.
Hence, \(m\) is a trap space of \(\mathcal{N}\).
In addition, since \(\left\vert S_{\mathcal{N}}[m]\right\vert = 1\), \(m\) is also a minimal trap space.
To compute all fixed points of \(\mathcal{N}\), we can add more constraints to the encoded ASP characterizing all conflict-free siphons (equivalently trap spaces).
For every \(v \in V\), we add to the encoded ASP the rule
\[
  \verb|p-v ; n-v.|
\]
that ensures that for every conflict-free siphon \(S\), it contains either \(\verb|p-v|\) or \(\verb|n-v|\) for every \(v \in V\).
Equivalently, the trap space corresponding to \(S\) is always a fixed point.
Now, the set of answer sets of the encoded ASP is equivalent to the set of fixed points of \(\mathcal{N}\).
In particular, when solving the encoded ASP using an ASP solver, we do not need to use the built-in option for computing set-inclusion maximal answer sets.
Note that we can also build another ASP characterizing all fixed points of \(\mathcal{N}\) based on the equivalence between a fixed point of \(\mathcal{N}\) and a deadlock of its Petri net encoding~\cite{chatain2014characterization}.
This approach may give a more compact ASP.

Third, we consider the trap spaces intersecting a given subspace \(m^{*}\) of a Boolean network.
A trap space \(m\) intersects \(m^{*}\) if and only if \(S_{\mathcal{N}}[m] \cap S_{\mathcal{N}}[m^{*}] \neq \emptyset\).
It follows that for every \(v\), if \(m^{*}(v) = 0\) then \(m(v) = 0\) or \(m(v) = \star\), if \(m^{*}(v) = 1\) then \(m(v) = 1\) or \(m(v) = \star\).
For the former case, we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|:- n-v.|
\]
that ensures that \(m(v)\) cannot be 1.
For the latter case, we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|:- p-v.|
\]
that ensures that \(m(v)\) cannot be 0.
Now \(\mathcal{L}\) characterizes all trap spaces that intersect \(m^{*}\).

Finally, we consider the trap spaces that are inside a given subspace \(m^{*}\) of a Boolean network.
We first adjust \(\mathcal{L}\) to characterize all such trap spaces.
A trap space \(m\) is inside \(m^{*}\) if and only if \(m(v) = m^{*}(v)\) for every \(v \in D_m\).
If \(m^{*}(v) = 0\), we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|p-v.|
\]
that ensures that \(m(v) = 0\).
If \(m^{*}(v) = 1\), we add to \(\mathcal{L}\) the ASP rule
\[
  \verb|n-v.|
\]
that ensures that \(m(v) = 1\).
It is noted that if we want to compute maximal trap spaces inside \(m^{*}\), we need to exclude the conflict-free siphon corresponding \(m^{*}\) from the solution space.
Specifically, we need to add to \(\mathcal{L}\) the ASP rule
\[
  \verb|p-v_i1| ; \verb|n-v_i1| ; \dots ; \verb|p-v_ik| ; \verb|n-v_ik|.
\]
where \(\{v_{i_1}, \dots, v_{i_k}\}\) is the set of free nodes of \(m^{*}\).
This rule ensures that \(m \neq m^{*}\).
In the case that \(m^{*} = \varepsilon\), we have all maximal trap spaces of the original Boolean network.
